Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-08-05 resubmission/blinded-paper-resubmit.tex
Update + render
Creo que solo falta mirar en detalle la conclusión @juancarloscastillo y luego lo cerramos con un vistazo final

\date{}
\begin{document}
\maketitle
\begin{abstract}
Several countries have experienced a shift toward the privatization and
commodification of public goods, welfare policies, and social services.
In Latin America, Chile stands out as a paradigmatic case where this
trend has led to the extensive marketization of essential services. From
a moral economy perspective, the extent to which individuals consider it
fair for access to such services to depend on market criteria has been
conceptualized as market justice preferences. This study investigates
the relationship between perceptions of economic inequality,
meritocratic beliefs, and market justice preferences in Chile between
2016 and 2023. Using six waves of panel data from the Chilean
Longitudinal Social Survey -- ELSOC (\(N_{observations}\) = 8,643;
\(N_{individuals}\) = 1,687), the analysis examines how subjective
assessments of inequality shape attitudes toward the role of merit in
access to key social services such as healthcare, education, and
pensions. The findings show that greater perceived inequality is
associated with lower market justice preferences. However, individuals
who believe that effort is rewarded are more likely to legitimize
existing disparities. In contrast, the perception that talent is
rewarded shows a negative effect on market justice preferences; an
effect that intensifies as perceived inequality increases over time. The
study also considers the influence of major social movements during this
period, which appear to have reshaped public discourse on justice and
fairness. These findings contribute to a deeper understanding of how
distributive beliefs evolve in contexts marked by persistent inequality
and entrenched neoliberal frameworks \newline \textbf{Keywords}:
Economic inequality, meritocracy, market justice, Chile, public
preferences, inequality perception
\end{abstract}

@@

Since 1980, economic inequality and wealth concentration have
dramatically increased worldwide, becoming one of the main challenges
for the social sciences. Globally, in 2021, less than 50\% of the world
population owned only 2\% of the wealth, while the richest 10\%
concentrated 76\%. Futhermore, the wealthiest 1\% captured nearly 38\%
of total assets (\citeproc{ref-chancel_world_2022}{Chancel et al.,
2022}). This context of economic disparity has sparked renewed interest
in studying not only the objective aspects of inequality, such as income
and access to resources, but also its subjective dimensions, including
perceptions, beliefs, and associated attitudes
(\citeproc{ref-janmaat_subjective_2013}{Janmaat, 2013}). Such dimensions
are deemed as key to understand the political delegitimation that
economic inequality can elicit
(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022}).
Therefore, understanding how people perceive inequality is crucial, as
these perceptions can influence how societies comprehend and justify (or
not) inequality in the distribution of goods and services, as well as
their capacity to funnel political discontent and eventually opposition
to the current economic distribution.

The justification of economic inequality has a long tradition in
sociology, with seminal works such as the ones from Kluegel
(\citeproc{ref-kluegel_social_1995a}{1995};
\citeproc{ref-kluegel_beliefs_1986}{1986}) and Lane
(\citeproc{ref-lane_political_1962}{1962},
\citeproc{ref-lane_market_1986}{1986}). Lane's main aim was to ``help
explain the tenacious hold of market capitalism on the public mind''
(\citeproc{ref-lane_market_1986}{1986, p. 384}), proposing the
distinction between market and political justice. Whereas political
justice refer to government actions aligned with justice principles such
equality or need, market justice involves the allocation linked to
earned desserts, particularly through individual contributions based on
as effort and ability. Sociological survey research in this area has
considered questions related to the extent to which it is considered
just to get better access to social services (such as pensions, health
and education) based on payment capacity
(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}). Although
these services could be considered basic rights of citizens, evidence to
date suggests that a significant portion of the population justifies
inequality in these areas (\citeproc{ref-koos_moral_2019}{Koos \&
Sachweh, 2019}) and that such preferences vary according to objective
socioeconomic factors and the contextual characteristics of countries
(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In the present paper, we
argue that there are two other relevant factors to take into account
when analyzing differences between individuals and groups to this
regard: the perception of economic inequality and the perception of
meritocracy.

The perception of economic inequality can be understood as an
individual's subjective assessment of how resources are allocated among
members of a given society (\citeproc{ref-akyelken_urban_2020}{Akyelken,
2020}). Regardless of their measurement, various studies have shown that
this perception often underestimates the gap between the rich and the
poor, revealing inconsistencies between objective and subjective aspects
of inequality (\citeproc{ref-trump_income_2018}{Trump, 2018}). Moreover,
there is consistent evidence that perceived inequality, even under
control of objective inequality, could have implications for attitudes
toward the distribution of goods and services
(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
\citeproc{ref-schroder_income_2017}{Schröder, 2017}), redistributive
preferences, justification of inequality, and legitimacy of the economic
system (\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
\citeproc{ref-garcia-sanchez_attitudes_2020}{García-Sánchez et al.,
2020}, \citeproc{ref-garcia-sanchez_perceptions_2018}{2018}). From this
perspective, in this paper we argue that perceptions of economic
inequality may play a relevant aspect for understanding market justice
preferences, as a lower perceived inequality could reinforce the idea
that the differences in access to social services could be justified ---
a link that becomes difficult to hold when high inequality is perceived.

Along with perception of inequality, a second focus of this paper is the
perception of meritocracy. Meritocracy posits that inequality can be
legitimized through distributive criteria such as effort and talent
(\citeproc{ref-davis_principles_2001}{Davis \& Moore, 2001};
\citeproc{ref-young_rise_1962}{Young, 1962}). Previous studies have
demonstrated that people with stronger meritocratic beliefs tend to
perceive less inequality as they attribute economic differences to
individual achievements (\citeproc{ref-mijs_paradox_2021}{Mijs, 2021};
\citeproc{ref-wilson_role_2003}{Wilson, 2003}) and also to justify more
inequality as these beliefs are associated with attitudes that
legitimize status differences
(\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023}). In a context
as Chile, where the distribution of goods and services is predominantly
governed by market logics strongly introduced during the military
dictatorship (1973-1989) (\citeproc{ref-boccardo_30_2020}{Boccardo,
2020}; \citeproc{ref-madariaga_three_2020}{Madariaga, 2020}),
meritocratic perceptions have the potential to affect the acceptance of
social inequalities. Therefore, incorporating the perception of
meritocracy in this literature allows for an understanding of how
attitudes toward inequality are shaped not only by the perception of
economic gaps, but also by beliefs about how these gaps are justified.
Furthermore, these beliefs are often consolidated from an early age,
reinforced by institutions promoting values such as effort and
individual skills to climb socially
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-reynolds_perceptions_2014}{Reynolds \& Xian, 2014}).

At this point we can propose the following research questions: To what
extent the preferences for market justice are associated with
perceptions of inequality and meritocracy? How stable are these
associations (within subjects) along time? The primary objective of this
study is to analyze the interplay between perceptions of inequality and
meritocracy and their joint influence on preferences for market justice
along time in Chile using the panel survey ELSOC (Social Chilean
Longitudinal Study, 2016-2023). The chilean case appears as a particular
case for studying market justice preferences given their story of high
inequality and strong free-market policies introduced in the
dictatorship period (\citeproc{ref-flores_top_2020}{Flores et al.,
2020}; \citeproc{ref-madariaga_three_2020}{Madariaga, 2020}).
Additionally, this analysis seeks to explore how political and social
contingencies---such as the 2019 and 2022 social movements---might have
moderated these relationships by prompting more critical reflection on
the commodification of social services
(\citeproc{ref-somma_no_2021}{Somma et al., 2021}). Recognizing the
temporal dimension in shaping market justice preferences is essential,
given that historical and contextual factors that could challenge or
reaffirm the way in which individuals persceive inequality and
meritocracy, and consequantely --- following our argument --- the
preferences for market justice.

\section{Theoretical views on market justice, inequality perception, and
meritocracy}\label{theoretical-views-on-market-justice-inequality-perception-and-meritocracy}

\subsection{The justification of market
inequality}\label{the-justification-of-market-inequality}

Conceptually, \emph{market justice} has been discussed in the literature
as a normative principle that legitimates the distribution of economic
rewards based on individual merit. It is possible to trace the concept
to the understanding of Lane (\citeproc{ref-lane_market_1986}{1986}),
who makes a contrast between market justice and political justice. The
author defines market justice as a system of ``earned deserts'', whereby
individuals are seen as deserving of a determined distributive outcome
due to their effort and skills. In contrast, political justice
emphasizes principles of equality and need, which are often represented
by the welfare state action through social policies. An important remark
is that the principles of market justice rely on the assumption that
markets are neutral and self-regulating arenas, where individuals are
treated fairly because they face the same formal rules of engagement
(\citeproc{ref-lane_market_1986}{Lane, 1986}). Consequently, the
legitimacy of market justice stems from the assumption that inequalities
are not only inevitable but fair---so long as the rules are transparent
and opportunities are open. In this way, market justice provides a moral
justification for inequality by framing it as a necessary outcome of
individual responsibility
(\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}).

Empirical studies have shown different strategies for the study of
market justice preferences. A common approach in the literature is to
gauge attitudes toward the legitimacy of inequality in specific domains,
especially when linked to income differences. This can be traced to the
seminal work of Kluegel and Smith
(\citeproc{ref-kluegel_beliefs_1981}{1981}) who assessed the normative
foundations that explain public support for economic inequality. Over
time, this approach has been extended beyond income to include other
market-mediated outcomes, such as education, healthcare, and/or
pensions. For example, Von Dem Knesebeck et al.
(\citeproc{ref-vondemknesebeck_are_2016}{2016}) and Immergut and
Schneider (\citeproc{ref-immergut_it_2020}{2020}) examine whether
citizens consider it fair that individuals with higher incomes can
access better healthcare, while Lee and Stacey
(\citeproc{ref-lee_fairness_2023}{2023}) apply a similar method in the
context of education in Australia. These studies usually rely on a
survey item asking respondents to evaluate the fairness of income-based
access to welfare services, allowing for comparing justice perceptions
across different contexts. Similarly, comparative studies have also
considered these indicators to study the support for market-based
distribution in welfare systems (\citeproc{ref-lindh_public_2015}{Lindh,
2015}; \citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). More
recently, Castillo et al.
(\citeproc{ref-castillo_socialization_2024}{2024}) introduced a
single-item composite measure of market justice to assess student
attitudes toward income-based access to education, healthcare, and
pensions in Chile. These empirical strategies all aim to capture the
extent to which individuals accept inequality when framed as a
reflection of market outcomes.

The study of market justice preferences has increasingly focused on how
they are shaped by individuals' socioeconomic position, normative
beliefs, and the institutional context in which they are embedded.
Across the literature, there is empirical evidence suggesting that
individuals in more advantaged socioeconomic positions---those with
higher occupational class, income, and education---are more likely to
support market justice principles (\citeproc{ref-koos_moral_2019}{Koos
\& Sachweh, 2019}; \citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). This tendency
reflects not only material self-interest but also a broader moral
economy, in which winners of the market system internalize
justifications for the status quo. At the same time, political ideology
also plays a role ---such as economically conservative values --- where
more right-wing individuals show higher support for meritocracy and more
skepticism towards redistribution. This is particularly salient in
countries with more restricted public provision of social services. For
example, Castillo et al.
(\citeproc{ref-castillo_socialization_2024}{2024}) in Chile and, Lee and
Stacey (\citeproc{ref-lee_fairness_2023}{2023}) in Australia show that
right-wing individuals are more supportive of market-based distribution
of welfare. Beyond individual characteristics, country-level
institutions also play a central role. In liberal welfare regimes like
those of the United States or the United Kingdom, market justice
preferences are more widespread, while in coordinated or
social-democratic regimes---such as Sweden or Germany---citizens are
generally more critical of market-based inequalities
(\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). These findings suggest
that market justice is also shaped contextually, which is represented by
the development of welfare institutions, in line with the
policy-feedbacks literature
(\citeproc{ref-campbell_policy_2012}{Campbell, 2012};
\citeproc{ref-pierson_when_1993}{Pierson, 1993}).

\subsection{Perception of inequality}\label{perception-of-inequality}

Perceptions of inequality have been associated with a range of
distributive attitudes (\citeproc{ref-castillo_perception_2022}{Castillo
et al., 2022}; \citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez
et al., 2019}; \citeproc{ref-mccall_exposure_2017}{McCall et al., 2017};
\citeproc{ref-trump_income_2018}{Trump, 2018}). Research indicates that
perceptions of inequality influence redistributive preferences to a
greater extent than objective or actual economic inequality
(\citeproc{ref-gimpelson_misperceiving_2018}{Gimpelson \& Treisman,
2018}). Others have shown that the perception of inequality, through the
perceived wage gap, is related to the justification of greater
inequality (\citeproc{ref-Castillo2011}{Castillo, 2011}). In that sense,
lower perceived inequality can reinforce support for market-based
distributive arrangements by suggesting that the system is fair and that
outcomes reflect effort and ability (\citeproc{ref-kuhn_eye_2011}{Kuhn,
2011}). In contrast, when inequality is perceived as excessive or
structurally determined, individuals could question the legitimacy of
market justice and become more supportive of redistributive policies
(\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
2019}).

@@

Among the different approaches to measuring perceived inequality, one of
the most widely used is the estimation of wage gaps between occupational
extremes, such as between a CEO and a manual worker. This type of item
provides a concrete frame that respondents can relate to more easily
than abstract questions about national income distribution
(\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012};
\citeproc{ref-easterbrook_social_2021}{Easterbrook, 2021};
\citeproc{ref-willis_legitimacy_2015}{Willis et al., 2015}). While it
enables the estimation of inequality using simple heuristics, this
method is not without challenges. For instance, people often lack
reliable knowledge about the earnings of those at the top of the income
ladder, which leads to high variability in responses and the use of
biased mental shortcuts (\citeproc{ref-knell_perceptions_2020}{Knell \&
Stix, 2020}). Despite this, perceived wage gaps are strong predictors of
political attitudes
(\citeproc{ref-garcia-sanchez_perceptions_2018}{García-Sánchez et al.,
2018}; \citeproc{ref-pedersen_attitudes_2019}{Pedersen \& Mutz, 2019}),
making them a valuable tool for understanding public responses to
economic disparities.

Another relevant critique of this approach lies in its conflation of
different psychological constructs. Many surveys assess perceived
inequality through Likert-type items that ask respondents to agree or
disagree with statements, such as ``income differences are too large,''
which captures general concern or discomfort rather than a specific
perception (\citeproc{ref-Castillo2011}{Castillo, 2011};
\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
2019}). These items mix cognitive estimations with affective
evaluations, complicating the interpretation of what respondents
perceive versus what they morally reject. As a result, the conceptual
clarity between perceived inequality and inequality aversion remains
blurred in many empirical studies. To address this limitation, recent
work has emphasized the need to distinguish between absolute and
comparative measures, as well as between ideal and actual estimates of
economic gaps (\citeproc{ref-auspurg_why_2017}{Auspurg et al., 2017};
\citeproc{ref-garcia-sanchez_creencias_2022}{García-Sánchez \& De
Carvalho, 2022}).

\subsection{Perception of meritocracy}\label{perception-of-meritocracy}

@@

Chile offers an interesting case for examining how public attitudes
toward the distribution of social services evolve amid declining poverty
and persistently high income inequality within a residual welfare
framework (\citeproc{ref-ferre_welfare_2023}{Ferre, 2023}). Following
the neoliberal reforms of the 1980s, the Chilean welfare state has
become increasingly reliant on private provision, with essential
services often commodified and largely reserved for those with
sufficient purchasing power (\citeproc{ref-boccardo_30_2020}{Boccardo,
2020}; \citeproc{ref-madariaga_three_2020}{Madariaga, 2020}). This model
has favored the affluent, while lower-income groups are left dependent
on limited and underfunded public alternatives. Research on social
stratification characterizes Chile as a society where mobility into
intermediate classes is feasible, but access to elite strata remains
largely restricted
(\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
2021}; \citeproc{ref-torche_intergenerational_2014a}{Torche, 2014}).
While some indicators---such as education and occupation---suggest a
relatively fluid class structure, income mobility remains constrained
(\citeproc{ref-espinoza_movilidad_2014}{Espinoza \& Núñez, 2014}).

@@

The survey has been conducted every year since 2016, except in 2020,
when it was suspended due to the COVID-19 pandemic. This study uses six
waves: 2016, 2017, 2018, 2019, 2022, and 2023. The 2021 wave was
excluded because a reduced version of the questionnaire omitted key
variables of interest. Between waves 1 and 6, panel attrition reached
40\%, resulting in a final two-level sample comprising N = 8,643
observations nested within N = 1,687 individuals. Longitudinal weights
are applied to adjust for both the sampling design and potential biases
from systematic non-response. Further details on sampling, attrition,
and weighting procedures are available at
https://coes.cl/encuesta-panel/, and the dataset is publicly accessible
at https://dataverse.harvard.edu/dataverse/elsoc.

@@

\bibitem[\citeproctext]{ref-campbell_policy_2012}
Campbell, A. L. (2012). Policy {Makes Mass Politics}. \emph{Annual
Review of Political Science}, \emph{15}(Volume 15, 2012), 333--351.
\url{https://doi.org/10.1146/annurev-polisci-012610-135202}

@@

\bibitem[\citeproctext]{ref-knell_perceptions_2020}
Knell, M., \& Stix, H. (2020). Perceptions of inequality. \emph{European
Journal of Political Economy}, \emph{65}, 101927.
\url{https://doi.org/10.1016/j.ejpoleco.2020.101927}

@@

\bibitem[\citeproctext]{ref-reynolds_perceptions_2014}
Reynolds, J., \& Xian, H. (2014). Perceptions of meritocracy in the land
of opportunity. \emph{Research in Social Stratification and Mobility},
\emph{36}, 121--137. \url{https://doi.org/10.1016/j.rssm.2014.03.001}

>>>>

\date{}
\begin{document}
\maketitle
\begin{abstract}
Several countries have experienced a shift toward the privatization and
commodification of public goods, welfare policies, and social services.
In Latin America, Chile stands out as a paradigmatic case where this
trend has led to the extensive marketization of essential services. From
a moral economy perspective, the extent to which individuals consider it
fair for access to such services to depend on market criteria has been
conceptualized as market justice preferences. This study investigates
the relationship between perceptions of economic inequality,
meritocratic beliefs, and market justice preferences in Chile between
2016 and 2023. Using six waves of panel data from the Chilean
Longitudinal Social Survey -- ELSOC (\(N_{observations}\) = 8,643;
\(N_{individuals}\) = 1,687), the analysis examines how subjective
assessments of inequality shape attitudes toward the role of merit in
access to key social services such as healthcare, education, and
pensions. The findings show that greater perceived inequality is
associated with lower market justice preferences. However, individuals
who believe that effort is rewarded are more likely to legitimize
existing disparities. In contrast, the perception that talent is
rewarded shows a negative effect on market justice preferences; an
effect that intensifies as perceived inequality increases over time. The
study also considers the influence of major social movements during this
period, which appear to have reshaped public discourse on justice and
fairness. These findings contribute to a deeper understanding of how
distributive beliefs evolve in contexts marked by persistent inequality
and entrenched neoliberal frameworks \newline \textbf{Keywords}:
Inequality perception, meritocracy, commodification, Chile, public
preferences
\end{abstract}

@@

Since 1980, economic inequality and wealth concentration have
dramatically increased worldwide, becoming one of the main challenges
for the social sciences. Globally, in 2021, less than 50\% of the world
population owned only 2\% of the wealth, while the richest 10\%
concentrated 76\%. Futhermore, the wealthiest 1\% captured nearly 38\%
of total assets (\citeproc{ref-chancel_world_2022}{Chancel et al.,
2022}). This context of economic disparity has sparked renewed interest
in studying not only the objective aspects of inequality, such as income
and access to resources, but also its subjective dimensions, including
perceptions, beliefs, and associated attitudes
(\citeproc{ref-janmaat_subjective_2013}{Janmaat, 2013}). Such dimensions
are deemed as key to understand the delegitimation of political
institutions that economic inequality can elicit
(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022}).
Therefore, understanding how people perceive inequality is crucial, as
these perceptions can influence how societies comprehend and justify (or
not) inequality in the distribution of goods and services, as well as
their capacity to funnel political discontent and eventually opposition
to the current economic distribution.

The justification of economic inequality is a topic with a long
tradition in sociology, with seminal works such as the ones from Kluegel
(\citeproc{ref-kluegel_social_1995a}{1995};
\citeproc{ref-kluegel_beliefs_1986}{1986}) and Lane
(\citeproc{ref-lane_political_1962}{1962},
\citeproc{ref-lane_market_1986}{1986}). Lane's main aim was to ``help
explain the tenacious hold of market capitalism on the public mind''
(\citeproc{ref-lane_market_1986}{1986, p. 384}), proposing the
distinction between market and political justice. Whereas political
justice refer to government actions aligned with justice principles such
equality or need, market justice involves the allocation linked to
earned desserts, particularly through individual contributions based on
individual merit. Part of this agenda has focused on analyzing the
extent to which individuals consider as just to get better access to
social services (such as pensions, health and education) based on
payment capacity (\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}), usually
under the label of market justice preferences. Although these services
can be considered fundamental rights of citizenship, existing evidence
indicates that a substantial share of the population legitimizes
inequality in these domains, and that such preferences are shaped by
individuals' socioeconomic characteristics as well as the broader
contextual features of their countries
(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). Within this framework,
we argue that there are two relevant factors that are usually sidelined
in this literature and that are key to understand market justice
preferences: the perception of economic inequality and the perception of
meritocracy.

The perception of economic inequality can be understood as an
individual's subjective assessment of how resources are allocated among
members of a given society (\citeproc{ref-akyelken_urban_2020}{Akyelken,
2020}). Regardless of their measurement, various studies have shown that
this perception often underestimates the gap between the rich and the
poor, revealing inconsistencies between objective and subjective aspects
of inequality (\citeproc{ref-trump_income_2018}{Trump, 2018}). Moreover,
there is consistent evidence that perceived inequality, even under
control of objective inequality, could have implications for attitudes
toward the distribution of goods and services
(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
\citeproc{ref-schroder_income_2017}{Schröder, 2017}), redistributive
preferences, justification of inequality, and legitimacy of the economic
system (\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
\citeproc{ref-garcia-sanchez_attitudes_2020}{García-Sánchez et al.,
2020}, \citeproc{ref-garcia-sanchez_perceptions_2018}{2018}). From this
perspective, in this paper we argue that perceptions of economic
inequality would play a relevant role for understanding market justice
preferences, as a lower perceived inequality could reinforce the idea
that the differences in access to social services could be justified ---
a link that becomes harder to hold when high inequality is perceived.

Along with perception of inequality, a second focus of this paper is the
perception of meritocracy. Meritocracy posits that inequality can be
legitimized through distributive criteria such as effort and talent
(\citeproc{ref-davis_principles_2001}{Davis \& Moore, 2001};
\citeproc{ref-young_rise_1962}{Young, 1962}). Previous studies have
shown that those with stronger meritocratic perceptions tend to justify
larger inequality, as economic differences are legitimized by being
attributed to individual achievements
(\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023};
\citeproc{ref-mijs_paradox_2021}{Mijs, 2021};
\citeproc{ref-wilson_role_2003}{Wilson, 2003}). We argue that in
contexts where the distribution of goods and services is predominantly
governed by market logics, as in the chilean case
(\citeproc{ref-boccardo_30_2020}{Boccardo, 2020};
\citeproc{ref-madariaga_three_2020}{Madariaga, 2020}), meritocratic
perceptions could play a role in the preferences of market justice.

The primary objective of this study is to analyze the interplay between
perceptions of inequality and meritocracy and their joint influence on
preferences for market justice along time in Chile using the panel
survey ELSOC (Social Chilean Longitudinal Study, 2016-2023).
Furthermore, we explore how political and social contingencies---such as
the 2019 and 2022 social movements---might have moderated these
relationships by prompting more critical reflection on the
commodification of social services (\citeproc{ref-somma_no_2021}{Somma
et al., 2021}). Recognizing the temporal dimension in shaping
preferences is essential, given that historical and contextual factors
that could challenge or reaffirm the way in which individuals perceive
inequality and meritocracy, resulting in changes in market justice
preferences.

\section{Market justice preferences}\label{market-justice-preferences}

Conceptually, \emph{market justice} has been discussed in the literature
as a normative principle that legitimates the distribution of economic
rewards based on individual merit. It is possible to trace the concept
to the understanding of Lane (\citeproc{ref-lane_market_1986}{1986}),
who makes a contrast between market justice and political justice. The
author defines market justice as a system of ``earned deserts'', whereby
individuals are seen as deserving of a determined distributive outcome
due to their effort and skills. In contrast, political justice
emphasizes principles of equality and need, which are often represented
by the welfare state action through social policies. An important remark
is that the principles of market justice rely on the assumption that
markets are neutral and self-regulating arenas, where individuals are
treated fairly because they face the same formal rules of engagement and
procedures (\citeproc{ref-lane_market_1986}{Lane, 1986}). Consequently,
the legitimacy of market justice stems from the assumption that
inequalities are not only inevitable but fair---so long as the rules are
transparent and opportunities are open. In this way, market justice
provides a moral justification for inequality by framing it as a
necessary outcome of individual responsibility
(\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}).

Empirical studies have shown different strategies for the study of
market justice preferences. A common approach in the literature is to
gauge attitudes toward the legitimacy of inequality in specific domains,
especially when linked to income differences. This can be traced to the
seminal work of Kluegel and Smith
(\citeproc{ref-kluegel_beliefs_1981}{1981}) who assessed the normative
foundations that explain public support for economic inequality. Over
time, this approach has been extended beyond income to include other
market-mediated outcomes, such as education, healthcare, and/or
pensions. For example, Von Dem Knesebeck et al.
(\citeproc{ref-vondemknesebeck_are_2016}{2016}) and Immergut and
Schneider (\citeproc{ref-immergut_it_2020}{2020}) examine whether
citizens consider it fair that individuals with higher incomes can
access better healthcare, while Lee and Stacey
(\citeproc{ref-lee_fairness_2023}{2023}) apply a similar method in the
context of education in Australia. These studies usually rely on survey
items asking respondents to evaluate the fairness of income-based access
to welfare and social services(\citeproc{ref-lindh_public_2015}{Lindh,
2015}; \citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). More
recently, Castillo et al.
(\citeproc{ref-castillo_socialization_2024}{2024}) introduced a
single-item composite measure of market justice to assess student
attitudes toward income-based access to education, healthcare, and
pensions in Chile. Such empirical strategies aim to capture the extent
to which individuals accept inequality when framed as a reflection of
market outcomes.

The study of market justice preferences has increasingly focused on how
they are shaped by individuals' socioeconomic position, normative
beliefs, and the institutional context in which they are embedded
(\citeproc{ref-pierson_when_1993}{Pierson, 1993}). Across the
literature, there is empirical evidence suggesting that individuals in
more advantaged socioeconomic positions---those with higher occupational
class, income, and education---are more likely to support market justice
principles (\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). This tendency
reflects not only material self-interest but also a broader moral
economy, in which winners of the market system internalize
justifications for the status quo. At the same time, political ideology
also plays a role ---such as economically conservative values --- where
right-wing individuals show higher support for meritocracy and more
skepticism towards redistribution. This is particularly salient in
countries with more restricted public provision of social services. For
example, in cases such as Chile
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024}) and
Australia(\citeproc{ref-lee_fairness_2023}{Lee \& Stacey, 2023}) the
evidence suggests that right-wing individuals are more supportive of
market-based distribution of welfare. Regarding contextual factors, in
liberal welfare regimes like those of the United States or the United
Kingdom, market justice preferences are more widespread, while in
coordinated or social-democratic regimes---such as Sweden or
Germany---citizens are generally more critical of market-based
inequalities (\citeproc{ref-immergut_it_2020}{Immergut \& Schneider,
2020}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}).

\section{Perceptions of inequality and
meritocracy}\label{perceptions-of-inequality-and-meritocracy}

\emph{The perception of economic inequality}

@@

Perceptions of inequality have been associated with a range of
distributive attitudes (\citeproc{ref-castillo_perception_2022}{Castillo
et al., 2022}; \citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez
et al., 2019}; \citeproc{ref-mccall_exposure_2017}{McCall et al., 2017};
\citeproc{ref-trump_income_2018}{Trump, 2018}). Research indicates that
perceptions of inequality influence redistributive preferences to a
greater extent than objective or actual economic inequality
(\citeproc{ref-Castillo2011}{Castillo, 2011};
\citeproc{ref-gimpelson_misperceiving_2018}{Gimpelson \& Treisman,
2018}). Lower perceived inequality can reinforce support for
market-based distributive arrangements
(\citeproc{ref-kuhn_eye_2011}{Kuhn, 2011}). In contrast, when inequality
is perceived as excessive, individuals could question the legitimacy of
market justice and become more supportive of redistributive policies
(\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
2019}).

Assessing the perception of inequality in empirical research is an
ongoing challenge. Specifically in survey research, the assessment of
perceived inequality through Likert-type items such as ``income
differences are too large'' have been questioned as they captures
general concern or discomfort rather than a specific perception
(\citeproc{ref-Castillo2011}{Castillo, 2011};
\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
2019}). As a result, the conceptual clarity between perceived inequality
and inequality aversion remains blurred in many empirical studies. To
address this limitation, recent work has emphasized the need to
distinguish between absolute and comparative measures, as well as
between ideal and perceived estimates of economic gaps
(\citeproc{ref-auspurg_why_2017}{Auspurg et al., 2017};
\citeproc{ref-garcia-sanchez_creencias_2022}{García-Sánchez \& De
Carvalho, 2022}). Through asking perceived salaries for occupations in
the extremes of the status continuum (such as a CEO and a manual
worker), an indirect measure of perceived inequality is built by the
ratio of the high and the low status occupation. This type of measure
provides a concrete frame that respondents can relate to more easily
than abstract questions about national income distribution
(\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012};
\citeproc{ref-easterbrook_social_2021}{Easterbrook, 2021};
\citeproc{ref-willis_legitimacy_2015}{Willis et al., 2015}). Evidence to
date shows that perceived wage gaps are strong predictors of political
attitudes (\citeproc{ref-garcia-sanchez_perceptions_2018}{García-Sánchez
et al., 2018}; \citeproc{ref-pedersen_attitudes_2019}{Pedersen \& Mutz,
2019}), making them a valuable tool for understanding public responses
to economic disparities.

\emph{Perception of meritocracy}

@@

Chile offers an interesting case for examining how public attitudes
toward the distribution of social services evolve amid declining poverty
and persistently high income inequality within a residual welfare
framework (\citeproc{ref-ferre_welfare_2023}{Ferre, 2023}). Following
the neoliberal reforms of the 1980s introduced during the civil-military
dictatorship (1973-1989), the Chilean welfare state has become
increasingly reliant on private provision, with essential services often
commodified and largely reserved for those with sufficient purchasing
power (\citeproc{ref-boccardo_30_2020}{Boccardo, 2020};
\citeproc{ref-madariaga_three_2020}{Madariaga, 2020}). This model has
favored the affluent, while lower-income groups are left dependent on
limited and underfunded public alternatives. Research on social
stratification characterizes Chile as a society where mobility into
intermediate classes is feasible, but access to elite strata remains
largely restricted
(\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
2021}; \citeproc{ref-torche_intergenerational_2014a}{Torche, 2014}).
While some indicators---such as education and occupation---suggest a
relatively fluid class structure, income mobility remains constrained
(\citeproc{ref-espinoza_movilidad_2014}{Espinoza \& Núñez, 2014}).

@@

The survey has been conducted every year since 2016, except in 2020,
when it was suspended due to the COVID-19 pandemic. This study uses six
waves: 2016, 2017, 2018, 2019, 2022, and 2023. The 2021 wave was
excluded because a reduced version of the questionnaire omitted key
variables of interest. Between waves 1 and 6, panel attrition reached
40\%, resulting in a final two-level sample comprising N = 8,643
observations nested within N = 1,687 individuals. Longitudinal weights
are applied to adjust for both the sampling design and potential biases
from systematic non-response. Further details on sampling, attrition,
and weighting procedures are available at
\url{https://coes.cl/encuesta-panel/}, and the dataset is publicly
accessible at \url{https://dataverse.harvard.edu/dataverse/elsoc}.

@@

@@

@@
